Sequential rank agreement methods for comparison of ranked lists

### 2015-04-21 latex/paper.Rnw

In this article we propose a method to measure similarity among ranked
lists for each depth in the list. As a measure for similarity we will
use the limits of agreement known from agreement between quantitative
variables \citep{Carstensen2011} but will compute the agreement on the
ranks of the items.  The proposed method allows for multiple lists
simultaneously, provides a dynamic measure of agreement for each depth
in the lists, places more weight on the top of the list, accommodates
censored/incomplete lists of varying lengths, and has a natural
interpretation that directly relates to the ranks. The general idea is
to define agreement based on the sequence of ranks from the first $k$
elements in each list. This approach allows us to estimate a potential
change point in the agreement in order to determine an optimal $k$
where there is a shift in the agreement among lists and enables us to
use randomization approaches to evaluate the rank agreement. In this
sense it is a combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches.

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$. An
ordered list is a permutation function, $R: \{1,\dots,P\}\to
\{1,\dots,P\}$, such that $R(X_p)$ is the rank of item $X_p$ in the
list. The inverse mapping $\pi=R^{-1}$ assigns to rank
$r\in\{1,\dots,P\}$ the item $\pi(r)$ found at that rank. The methods
described below work for a set of $L$ lists $R_1,\dots,R_L$,
$L\geq2$. We denote $\pi_l=R_l^{-1}$ for the corresponding inverse
mappings. Panels (a) and (b) of Table~\ref{tab:example} show a
schematic example of these mappings. Thus if $\pi_l(1)=X_{34}$ then
item $X_{34}$ is ranked first in list $l$.

@@

We analyzed these gene expression data using four different
approaches: marginal two-sample $t$ tests, marginal logistic
regression analyses, logistic regression eleastic net, and marginal
maximum information content correlations (MIC) \citep{}. For the first
two methods, the genes were ranked according to $p$ value.

>>>>

In this article we propose a method to measure similarity among ranked
lists for each depth in the list. As a measure for similarity we will
use the limits of agreement known from agreement between quantitative
variables \citep{Carstensen2010} but will compute the agreement on the
ranks of the items.  The proposed method allows for multiple lists
simultaneously, provides a dynamic measure of agreement for each depth
in the lists, places more weight on the top of the list, accommodates
censored/incomplete lists of varying lengths, and has a natural
interpretation that directly relates to the ranks. The general idea is
to define agreement based on the sequence of ranks from the first $k$
elements in each list. This approach allows us to estimate a potential
change point in the agreement in order to determine an optimal $k$
where there is a shift in the agreement among lists and enables us to
use randomization approaches to evaluate the rank agreement. In this
sense it is a combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches.

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$. An
ordered list is a permutation function, $R: \{1,\dots,P\}\to
\{1,\dots,P\}$, such that $R(X_p)$ is the rank of item $X_p$ in the
list. The inverse mapping $\pi=R^{-1}$ assigns to rank
$r\in\{1,\dots,P\}$ the item $\pi(r)$ found at that rank. The methods
described below work for a set of $L$ lists $R_1,\dots,R_L$,
$L\geq2$. We denote $\pi_l=R_l^{-1}$ for the corresponding inverse
mappings. Panels (a) and (b) of Table~\ref{tab:example} show a
schematic example of these mappings. Thus if $\pi_l(1)=X_{34}$ then
item $X_{34}$ is ranked first in list $l$, and similarly $R_l(X_{34})=1$.

@@

We analyzed these gene expression data using four different
approaches: marginal two-sample $t$ tests, marginal logistic
regression analyses, logistic regression eleastic net, and marginal
maximum information content correlations (MIC) \citep{Reshef2011}. For
the first two methods, the genes were ranked according to $p$ value.
